---
title: "MielleM hw002"
output: github_document
---
#### To start off, let's get tidyverse and gapminder loaded
Load gapminder data and tidyverse package
```{r}
library(tidyverse)
library(gapminder)
library(scales)
```


##Smell test the data

Question | Answer
---------|-------
Object type | data.frame
Class | tibble, or tbl_df
Columns | 6
Rows | 1704

Column | Type
-------|-----
country | factor
continent | factor
year | integer
lifeExp | double (numeric)
pop | integer
gdpPercap | double (numeric)

Most attributes of the tibble (e.g. number of observations) can be individually obtained by specific commands, rather than the larger summary created by glimpse(gapminder). If only one piece of information was needed, it would make more sense to target that specifically. 

The class command gives the type and class of gapminder. The glimpse command gives the number of observations, as well as the number, name, and data type of each variable. This is the neatest way to summarize what the tibble contains. 
```{r}
class(gapminder)
glimpse(gapminder)
```


##Explore individual variables


### Exploring a quantitative variable: GDP per capita  

#### With a five number summary:
```{r}
gdp <- gapminder::gapminder$gdpPercap

summary(gdp)
```
From the function above, we can see that GDP per capita ranges from $241.20 to $113,523.10, with a median value of $3,531.80.
The median and the mean ($7,215.30) are fairly different values, suggesting that this data is not normally distributed. 

#### Let's draw a kernel density plot check our assumption that the data is not normally distributed. 

```{r}
ggplot(gapminder, aes(gdpPercap)) +
  geom_density(color = "dark grey", fill = "light blue")+
  xlab("GDP per capita")
```

It appears that the data is not normally distributed, with the mode close to the median and a strong right skew. 


#### The same five number summary (minus the mean) from above, but drawn as a box plot
 
```{r}
boxplot(gdp)
```


#### The box plot makes it difficult to see much of a trend in the GDP data. Let's create a box plot for each continent so we can compare. 

```{r}
g <- ggplot(gapminder, aes(continent, gdpPercap))
```


```{r}
g +
  geom_boxplot() +
  scale_y_continuous(labels = comma) +
  xlab("Continent") +
  ylab("GDP per capita")
```

Let's try log-transforming GDP so the data doesn't appear so bunched at the bottom of the graph.
```{r}
g + 
  scale_y_log10() +
  geom_boxplot() +
  xlab("Continent") +
  ylab("GDP per capita")
```

Asia appears to have the widest range of GDP values, and Oceania has the smallest. Europe has a somewhat normal-looking distribution.  



### Exploring a qualitative variable: Continent  

#### First, let's make a basic histogram containing the data count for each continent. 

```{r}
c <- ggplot(gapminder, aes(continent))

c + 
  geom_bar()
```

This histogram looks pretty boring with all the grey. Plus, it doesn't help tell the story that I want to show with the graph. I'm trying to emphasize the differences between the continents, so I'd like to make them all different colors. 

#### First, we can check ColorBrewer for a list of available preset color palettes. 
```{r}
RColorBrewer::display.brewer.all()
```

#### Once I've selected the desired palette, I can use it to color the bar graphs. 
```{r}
c <- ggplot(gapminder, aes(continent, fill=continent))
c + 
  geom_bar() +
  geom_text(stat = "count", aes(label = ..count.., y = ..count..)) +
  scale_fill_brewer(palette = "Dark2")
```

In this histogram, I've also included the counts. It looks like Africa has significantly more data points than any other continent, the Americas, Asia, and Europe have a fairly similar number, and Oceania has significantly fewer associated observations. 

The two histograms above are for all the observations, from the 1950's onward. Because the number of countries in each continent has stayed relatively stable, a histogram for an individual year should look fairly similar to the histograms for all the data. Let's test this assumption with the most recent year that data is available. 

#### First, let's find the most recent data available. 

```{r}
tail(sort(gapminder$year), 1)
```


#### Then, let's filter the data to this year, and create a new histogram for 2007. 

```{r}
gapminder %>%
  filter(year == 2007) %>% 
  ggplot(aes(continent, fill = continent)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = ..count.., y = ..count..)) +
  scale_fill_brewer(palette = "Dark2")

```

As expected, this histogram is pretty similar to the one created with all the data. 



## Exploring various plot types


#### Here, I've created a line graph showing the progression of GDP over time. 

Each line represents a country, and are colored by continent. As you can see, it's pretty busy and therefor difficult to interpret. 

```{r}
v <- ggplot(gapminder, aes(year, gdpPercap, color = continent)) 

v + 
  geom_line(aes(group=country), alpha=0.3) +
  scale_color_brewer(palette = "Dark2") +
  theme_light()
```

To simplify, the graph below only displays data from countries with a population over 50,000,000. 

```{r}
gapminder %>%
  filter(pop > 50000000) %>% 
  ggplot(aes(year, gdpPercap, color = continent)) +
  geom_line(aes(group=country), alpha = 0.75) +
  scale_color_brewer(palette = "Dark2") +
  scale_y_continuous(labels = comma) +
  xlab("Year") +
  ylab("GDP per capita") +
  theme_light()
  
```

#### Visualizing three variables in one graph

I can visualize three variables by creating a plot displaying the distribution of GDP per capita by continent as both a violin plot and jitter plot. The dots of the jitter plot are colored to reflect life expectancy, with lower values represented by darker color. 


Here, I've filtered the years to 1980 and before. 
```{r}
gapminder %>%
  filter(year < 1980) %>% 
  ggplot(aes(continent, gdpPercap, color=lifeExp)) +
  geom_jitter() +
  geom_violin(alpha=0.6) +
  xlab("Continent") +
  ylab("GDP per capita") +
  theme_light()
```


Here is the same graph, but with years after 1980. Note that the legend and y axis scale is slightly different. 
```{r}
gapminder %>%
  filter(year > 1980) %>% 
  ggplot(aes(continent, gdpPercap, color=lifeExp)) +
  geom_jitter() +
  geom_violin(alpha=0.6) +
  xlab("Continent") +
  ylab("GDP per capita") +
  theme_light()
```

From these two graphs, we can see that there has been an increase in both life expectancy and GDP per capita in the years before and after 1980. 




## Extra goodies

> Q: Evaluate this code and describe the result. Presumably the analyst’s intent was to get the data for Rwanda and Afghanistan. Did they succeed? Why or why not? If not, what is the correct way to do this?

```{r}
filter(gapminder, country == c("Rwanda", "Afghanistan"))
```

A: If the analyst was trying to select both Rwanda and Afghanistan data, using this segment of code would not produce the desired result. This filter uses the logical function "=="" which is equivalent to the English word "and", and is only returning one of the two countries for each year that data exists. The correct function would be: 
```{r}
filter(gapminder, country %in% c("Rwanda", "Afghanistan"))
```

 "%in%" is equivalent to the English word "or", and returns all observations with "Rwanda" or "Afghanistan". 
   




























